---
title: "ggbipartite"
format: gfm
fig-format: retina
fig-dpi: 350
execute:
  echo: true
  warning: false
  message: false
---

ggbipartite builds inspectable ggplot2 layouts for bipartite networks and can
align either partition with a phylogenetic tree. Abundance ribbons and binary
presence-absence segments use the same validated input contract.

## Installation

Install the development version from GitHub:

```{r}
#| eval: false
pak::pak("Hide-Fun/ggbipartite")
```

## Quick start

Create a named interaction matrix, compute its layout once, and share that
layout across layers.

```{r}
#| label: quick-start
#| fig-path: man/figures/README-
#| fig-width: 7
#| fig-height: 5
#| out-width: 672px
library(ggplot2)
library(ggbipartite)

interaction_matrix <- matrix(
  c(
    4, 0, 1,
    1, 3, 0,
    0, 2, 5
  ),
  nrow = 3,
  byrow = TRUE,
  dimnames = list(
    c("host_a", "host_b", "host_c"),
    c("otu_a", "otu_b", "otu_c")
  )
)

network_layout <- layout_bipartite(interaction_matrix)

ggplot() +
  geom_bipnet_box(
    layout = network_layout,
    type = "row",
    fill = "#4477AA"
  ) +
  geom_bipnet_box(
    layout = network_layout,
    type = "column",
    fill = "#EE6677"
  ) +
  geom_bipnet_interaction(
    layout = network_layout,
    alpha = 0.45
  ) +
  coord_fixed() +
  theme_void()
```

`network_layout$nodes` and `network_layout$interactions` contain the exact data
drawn by the layers. Long data is also supported when `row`, `column`, and
`weight` are named explicitly.

For a standard tree-link-network composition, supply trees while creating the
layout and call `plot_bipartite()`.

## Independent Metadata Legends

Row and column metadata often describe different biological concepts. Keep
their scales separate when the same category name could mean different things
on each side of the network.

```{r}
#| label: independent-metadata-legends
#| fig-path: man/figures/README-
#| fig-width: 8
#| fig-height: 5
#| out-width: 768px
row_metadata <- tibble::tibble(
  host = rownames(interaction_matrix),
  family = c("Unknown", "Fagaceae", "Rosaceae")
)
column_metadata <- tibble::tibble(
  otu = colnames(interaction_matrix),
  guild = c("Unknown", "ECM", "AMF")
)

metadata_layout <- layout_bipartite(
  interaction_matrix,
  metadata_row = row_metadata,
  metadata_column = column_metadata,
  metadata_row_key = "host",
  metadata_column_key = "otu"
)

family_palette <- c(
  "Unknown" = "#9C755F",
  "Fagaceae" = "#4E79A7",
  "Rosaceae" = "#F28E2B"
)
guild_palette <- c(
  "Unknown" = "#BAB0AC",
  "ECM" = "#E15759",
  "AMF" = "#76B7B2"
)

ggplot() +
  geom_bipnet_interaction(
    layout = metadata_layout,
    mapping = aes(fill = column_guild),
    alpha = 0.45,
    show.legend = FALSE
  ) +
  geom_bipnet_box(
    layout = metadata_layout,
    type = "column",
    mapping = aes(fill = column_guild)
  ) +
  scale_fill_manual(
    name = "Symbiont guild",
    values = guild_palette,
    guide = guide_legend(order = 2)
  ) +
  ggnewscale::new_scale_fill() +
  geom_bipnet_box(
    layout = metadata_layout,
    type = "row",
    mapping = aes(fill = row_family)
  ) +
  scale_fill_manual(
    name = "Host family",
    values = family_palette,
    guide = guide_legend(order = 1)
  ) +
  coord_fixed() +
  theme_void()
```

`ggnewscale::new_scale_fill()` starts a second fill scale. This is different
from merely changing guide order: the two `"Unknown"` levels above are mapped
independently because they belong to different metadata variables.

## Abundance Networks With Trees

The same layout can carry abundance polygons, matched tree tips, connector
panels, styled tip labels, and internal-node support labels. The rendering code
for this publication-style composition is intentionally hidden here; it is more
verbose than the minimal API example, but it shows the figure type that
`ggbipartite` is designed to make practical.

```{r}
#| label: abundance-with-trees
#| echo: false
#| fig-path: man/figures/README-
#| fig-width: 14
#| fig-height: 5.2
#| out-width: 960px
set.seed(20260304)

host_phylo <- ape::rtree(10)
host_phylo$tip.label <- paste0("Host_", sprintf("%02d", 1:10))
host_phylo$node.label <- sample(
  70:100,
  host_phylo$Nnode,
  replace = TRUE
)

symbiont_phylo <- ape::rtree(6)
symbiont_phylo$tip.label <- paste0("OTU_", sprintf("%02d", 1:6))
symbiont_phylo$node.label <- sample(
  70:100,
  symbiont_phylo$Nnode,
  replace = TRUE
)

host_order <- get_tip_order(host_phylo) |> rev()
symbiont_order <- get_tip_order(symbiont_phylo) |> rev()

interaction_matrix_publishable <- matrix(
  c(
    120, 0, 40, 0, 15, 0,
    40, 2, 0, 0, 6, 1,
    0, 12, 0, 4, 0, 0,
    8, 0, 0, 22, 0, 5,
    0, 0, 5, 8, 18, 0,
    65, 0, 20, 0, 3, 0,
    0, 7, 2, 0, 0, 13,
    4, 0, 0, 30, 0, 0,
    0, 0, 8, 3, 25, 2,
    20, 1, 0, 0, 10, 0
  ),
  nrow = 10,
  byrow = TRUE,
  dimnames = list(
    paste0("Host_", sprintf("%02d", 1:10)),
    paste0("OTU_", sprintf("%02d", 1:6))
  )
)

interaction_matrix_publishable <- interaction_matrix_publishable[
  host_order,
  symbiont_order,
  drop = FALSE
]

metadata_row_publishable <- tibble::tibble(
  host = rownames(interaction_matrix_publishable),
  host_lineage = c(rep("Lineage_A", 6), rep("Lineage_B", 4))
)

metadata_column_publishable <- tibble::tibble(
  symbiont = colnames(interaction_matrix_publishable),
  guild = c("AMF", "AMF", "ECM", "ECM", "Saprotroph", "Unknown")
)

gap_abundance <- sum(interaction_matrix_publishable) / 12
link_gap_mm <- 3
label_family <- "Arial"
tip_text_size <- 3.5
node_text_size <- 3

measure_label_width_mm <- function(labels, family, size) {
  label_plot <- ggplot2::ggplot(
    tibble::tibble(labels = labels),
    ggplot2::aes(x = 0, y = 0, label = labels)
  ) +
    marquee::geom_marquee(family = family, size = size)
  label_grob <- ggplot2::layer_grob(label_plot)[[1L]]
  grid::convertWidth(
    grid::grobWidth(label_grob),
    "mm",
    valueOnly = TRUE
  )
}

layout_abundance <- layout_bipartite(
  interaction_matrix_publishable,
  interaction = "abundance",
  row_tree = host_phylo,
  column_tree = symbiont_phylo,
  metadata_row = metadata_row_publishable,
  metadata_column = metadata_column_publishable,
  metadata_row_key = "host",
  metadata_column_key = "symbiont",
  gap = gap_abundance
)

abundance_plot <- plot_bipartite(
  layout_abundance,
  widths = c(
    row_tree = 1.5,
    row_link = 0.28,
    network = 1.5,
    column_link = 0.28,
    column_tree = 1.5
  )
)

for (component_name in c("row_tree", "column_tree")) {
  tree_component <- abundance_plot$components[[component_name]]
  tree_component$layers <- Filter(
    function(layer) !inherits(layer$geom, "GeomBlank"),
    tree_component$layers
  )
  abundance_plot$components[[component_name]] <- tree_component
}

host_tree <- abundance_plot$components$row_tree
x_span_host_abundance <- diff(range(host_tree$data$x, na.rm = TRUE))
host_tree_margin <- host_tree$theme$plot.margin
host_tree_margin[2] <- grid::unit(
  measure_label_width_mm(
    host_tree$data$label[host_tree$data$isTip],
    family = label_family,
    size = tip_text_size
  ),
  "mm"
)

abundance_plot$components$row_tree <- host_tree +
  geom_tipmarquee(
    align = TRUE,
    hjust = 0,
    linetype = "dotted",
    linesize = 0.35,
    family = label_family,
    size = tip_text_size
  ) +
  geom_nodemarquee(
    mapping = ggplot2::aes(label = label),
    node = "internal",
    nudge_x = x_span_host_abundance * 0.05,
    nudge_y = diff(range(host_tree$data$y)) * 0.015,
    hjust = 0.5,
    family = label_family,
    size = node_text_size
  ) +
  ggplot2::scale_x_continuous(
    expand = ggplot2::expansion(mult = c(0.05, 0))
  ) +
  ggplot2::theme(plot.margin = host_tree_margin)

symbiont_tree <- abundance_plot$components$column_tree
x_span_symbiont_abundance <- diff(
  range(symbiont_tree$data$x, na.rm = TRUE)
)
symbiont_tree_margin <- symbiont_tree$theme$plot.margin
symbiont_tree_margin[4] <- grid::unit(
  measure_label_width_mm(
    symbiont_tree$data$label[symbiont_tree$data$isTip],
    family = label_family,
    size = tip_text_size
  ),
  "mm"
)

abundance_plot$components$column_tree <- symbiont_tree +
  geom_tipmarquee(
    align = TRUE,
    hjust = 1,
    linetype = "dotted",
    linesize = 0.35,
    family = label_family,
    size = tip_text_size
  ) +
  geom_nodemarquee(
    mapping = ggplot2::aes(label = label),
    node = "internal",
    nudge_x = -x_span_symbiont_abundance * 0.05,
    nudge_y = diff(range(symbiont_tree$data$y)) * 0.015,
    hjust = 0.5,
    family = label_family,
    size = node_text_size,
    colour = "grey20"
  ) +
  ggplot2::scale_x_reverse(
    expand = ggplot2::expansion(mult = c(0, 0.05))
  ) +
  ggplot2::theme(plot.margin = symbiont_tree_margin)

for (component_name in c("row_link", "column_link")) {
  link_plot <- abundance_plot$components[[component_name]]
  link_margin <- link_plot$theme$plot.margin
  link_margin[c(2, 4)] <- grid::unit(link_gap_mm, "mm")
  link_plot$layers <- list()
  abundance_plot$components[[component_name]] <- link_plot +
    ggplot2::geom_segment(
      linetype = "dotted",
      linewidth = 0.35,
      colour = "grey45"
    ) +
    ggplot2::geom_point(
      ggplot2::aes(x = x, y = y),
      size = 0.7,
      colour = "grey30"
    ) +
    ggplot2::geom_point(
      ggplot2::aes(x = xend, y = yend),
      size = 0.7,
      colour = "grey30"
    ) +
    ggplot2::theme(plot.margin = link_margin)
}

network_plot <- abundance_plot$components$network
network_plot$layers <- list()
abundance_plot$components$network <- network_plot +
  geom_bipnet_interaction(
    ggplot2::aes(fill = row_host_lineage),
    layout = layout_abundance,
    alpha = 0.45,
    linewidth = 0.1,
    colour = "grey35",
    show.legend = FALSE
  ) +
  geom_bipnet_box(
    ggplot2::aes(fill = row_host_lineage),
    layout = layout_abundance,
    type = "row",
    linewidth = 0.25,
    colour = "grey20"
  ) +
  geom_bipnet_box(
    layout = layout_abundance,
    type = "column",
    fill = "#223A5E",
    linewidth = 0.25,
    colour = "grey20"
  ) +
  ggplot2::scale_fill_manual(
    name = "Host lineage",
    values = c("Lineage_A" = "#4E79A7", "Lineage_B" = "#F28E2B")
  ) +
  ggplot2::theme(text = ggplot2::element_text(family = label_family))

as_patchwork(abundance_plot) +
  patchwork::plot_layout(guides = "collect") &
  ggplot2::theme(legend.position = "bottom")
```

## Documentation

- `vignette("getting-started", package = "ggbipartite")` covers matrix and long
  inputs, abundance ribbons, binary segments, and shared layers.
- `vignette("tree-integration", package = "ggbipartite")` covers strict tree ID
  matching, connector panels, and direct binary tip alignment.
- `vignette("layout-contract", package = "ggbipartite")` defines the stable
  `bipartite_layout` schema.
- `vignette("migration-v0.10.0", package = "ggbipartite")` lists behavior
  changes and replacements for 0.9.x workflows.

The Japanese guides in `vignettes/*-ja.qmd` remain supplementary source
documents; the English API and vignettes are the primary reference.

## Development

Bug reports and focused reproducible examples are welcome in the
[GitHub issue tracker](https://github.com/Hide-Fun/ggbipartite/issues).
