---
title: "ggbipartite"
format: gfm
execute:
  echo: true
  warning: false
  message: false
---

ggbipartite builds inspectable ggplot2 layouts for bipartite networks and can
align either partition with a phylogenetic tree. Abundance ribbons and binary
presence-absence segments use the same validated input contract.

## Installation

Install the development version from GitHub:

```{r}
#| eval: false
pak::pak("Hide-Fun/ggbipartite")
```

## Quick start

Create a named interaction matrix, compute its layout once, and share that
layout across layers.

```{r}
#| label: quick-start
#| fig-path: man/figures/README-
library(ggplot2)
library(ggbipartite)

interaction_matrix <- matrix(
  c(
    4, 0, 1,
    1, 3, 0,
    0, 2, 5
  ),
  nrow = 3,
  byrow = TRUE,
  dimnames = list(
    c("host_a", "host_b", "host_c"),
    c("otu_a", "otu_b", "otu_c")
  )
)

network_layout <- layout_bipartite(interaction_matrix)

ggplot() +
  geom_bipnet_box(
    layout = network_layout,
    type = "row",
    fill = "#4477AA"
  ) +
  geom_bipnet_box(
    layout = network_layout,
    type = "column",
    fill = "#EE6677"
  ) +
  geom_bipnet_interaction(
    layout = network_layout,
    alpha = 0.45
  ) +
  coord_fixed() +
  theme_void()
```

`network_layout$nodes` and `network_layout$interactions` contain the exact data
drawn by the layers. Long data is also supported when `row`, `column`, and
`weight` are named explicitly.

For a standard tree-link-network composition, supply trees while creating the
layout and call `plot_bipartite()`.

## Documentation

- `vignette("getting-started", package = "ggbipartite")` covers matrix and long
  inputs, abundance ribbons, binary segments, and shared layers.
- `vignette("tree-integration", package = "ggbipartite")` covers strict tree ID
  matching, connector panels, and direct binary tip alignment.
- `vignette("layout-contract", package = "ggbipartite")` defines the stable
  `bipartite_layout` schema.
- `vignette("migration-v0.10.0", package = "ggbipartite")` lists behavior
  changes and replacements for 0.9.x workflows.

The Japanese guides in `vignettes/*-ja.qmd` remain supplementary source
documents; the English API and vignettes are the primary reference.

## Development

Bug reports and focused reproducible examples are welcome in the
[GitHub issue tracker](https://github.com/Hide-Fun/ggbipartite/issues).
