---
title: "Getting started with ggbipartite"
format: html
execute:
  warning: false
  message: false
vignette: >
  %\VignetteIndexEntry{Getting started with ggbipartite}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

ggbipartite separates layout computation from drawing. Compute one reusable
layout, inspect its data, and pass it to any combination of layers.

```{r}
library(ggplot2)
library(ggbipartite)
```

## Matrix input

A matrix must have explicit row and column names.

```{r}
interaction_matrix <- matrix(
  c(
    4, 0, 1,
    1, 3, 0,
    0, 2, 5
  ),
  nrow = 3,
  byrow = TRUE,
  dimnames = list(
    c("host_a", "host_b", "host_c"),
    c("otu_a", "otu_b", "otu_c")
  )
)

network_layout <- layout_bipartite(interaction_matrix)
network_layout
```

The stable node and interaction tables can be inspected before plotting.

```{r}
network_layout$nodes
network_layout$interactions
```

The three layers below share those exact coordinates.

```{r}
#| fig-width: 7
#| fig-height: 5
ggplot() +
  geom_bipnet_box(
    layout = network_layout,
    type = "row",
    fill = "#4477AA"
  ) +
  geom_bipnet_box(
    layout = network_layout,
    type = "column",
    fill = "#EE6677"
  ) +
  geom_bipnet_interaction(
    layout = network_layout,
    alpha = 0.45
  ) +
  coord_fixed() +
  theme_void()
```

## Long input

Long data requires three explicit columns. Duplicate cells are rejected unless
you deliberately supply an aggregation function.

```{r}
interaction_data <- data.frame(
  host = c("host_a", "host_a", "host_b", "host_c", "host_c"),
  otu = c("otu_a", "otu_c", "otu_b", "otu_b", "otu_c"),
  abundance = c(4, 1, 3, 2, 5)
)

long_layout <- layout_bipartite(
  interaction_data,
  row = host,
  column = otu,
  weight = abundance
)
```

Set `interaction = "binary"` explicitly for presence-absence drawing. Positive
weights become one; zeros remain absent edges.

```{r}
binary_layout <- layout_bipartite(
  interaction_matrix,
  interaction = "binary"
)

ggplot() +
  geom_bipnet_point(layout = binary_layout, type = "row") +
  geom_bipnet_point(layout = binary_layout, type = "column") +
  geom_bipnet_interaction(layout = binary_layout) +
  coord_fixed() +
  theme_void()
```

See `vignette("tree-integration", package = "ggbipartite")` for tree matching
and the high-level composer.
