---
title: "Integrating phylogenetic trees"
format: html
fig-format: retina
fig-dpi: 350
execute:
  warning: false
  message: false
vignette: >
  %\VignetteIndexEntry{Integrating phylogenetic trees}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

Tree labels and data identifiers are compared exactly. A supplied tree also
defines the order of its network partition. The input trees are not rotated or
otherwise optimized, so topology and branch lengths remain under user control.

```{r}
library(ggbipartite)
```

```{r}
interaction_matrix <- matrix(
  c(
    3, 0, 1,
    1, 4, 0,
    0, 2, 5
  ),
  nrow = 3,
  byrow = TRUE,
  dimnames = list(
    c("host_a", "host_b", "host_c"),
    c("otu_a", "otu_b", "otu_c")
  )
)

row_tree <- ape::read.tree(
  text = "((host_a:1,host_b:1):1,host_c:2);"
)
column_tree <- ape::read.tree(
  text = "(otu_a:1,(otu_b:1,otu_c:1):1);"
)
```

`plot_bipartite()` returns a `ggbipartite_plot` with named, editable
`components`. Tree tips are unlabelled by default. The helper below adds the
identifiers in aligned columns, with dotted guides for tips that end before
the label column. Extra horizontal space keeps the labels inside their panels;
the tree coordinates and tip order are unchanged.

```{r}
add_tree_tip_labels <- function(tree_plot, hjust = 0) {
  stopifnot(inherits(tree_plot, "ggplot"), hjust %in% c(0, 1))
  tree_extent <- max(tree_plot$data$x)

  tree_plot +
    ggtree::geom_tiplab(
      align = TRUE,
      hjust = hjust,
      linetype = "dotted",
      linesize = 0.3,
      size = 3.5
    ) +
    ggplot2::expand_limits(x = 1.55 * tree_extent)
}
```

Abundance layouts use connector panels between tree tips and boxes. The
composer gives every panel the same vertical range and expansion, so each
connector endpoint shares the height of its corresponding tip or box centre.

```{r}
#| fig-width: 10
#| fig-height: 5
#| out-width: 960px
abundance_layout <- layout_bipartite(
  interaction_matrix,
  row_tree = row_tree,
  column_tree = column_tree
)
abundance_plot <- plot_bipartite(abundance_layout)
abundance_plot$components$row_tree <- add_tree_tip_labels(
  abundance_plot$components$row_tree
)
abundance_plot$components$column_tree <- add_tree_tip_labels(
  abundance_plot$components$column_tree,
  hjust = 1
)
as_patchwork(
  abundance_plot,
  widths = c(row_tree = 2, network = 3, column_tree = 2)
)
```

Binary layouts align node centres directly with tree tips and therefore do not
need connector panels. Panel widths can be changed without breaking this
vertical alignment. The column tree has a reversed horizontal axis, so its
labels use `hjust = 1` to face the network.

```{r}
#| fig-width: 9
#| fig-height: 5
#| out-width: 864px
binary_layout <- layout_bipartite(
  interaction_matrix,
  interaction = "binary",
  row_tree = row_tree,
  column_tree = column_tree
)
binary_plot <- plot_bipartite(binary_layout)
binary_plot$components$row_tree <- add_tree_tip_labels(
  binary_plot$components$row_tree
)
binary_plot$components$column_tree <- add_tree_tip_labels(
  binary_plot$components$column_tree,
  hjust = 1
)
as_patchwork(
  binary_plot,
  widths = c(row_tree = 2, network = 3.5, column_tree = 2)
)
```

By default, either side of an ID mismatch is an error and the message lists the
set differences. If discarding unmatched objects is scientifically justified,
opt in explicitly:

```{r}
#| eval: false
layout_bipartite(
  interaction_matrix,
  row_tree = row_tree,
  unmatched_data = "drop",
  unmatched_tree = "prune"
)
```

Component edits are applied before `as_patchwork()` or printing. The network
and component data remain available through `$layout`, so these presentation
changes do not require recomputing the layout.
