---
title: "Integrating phylogenetic trees"
format: html
execute:
  warning: false
  message: false
vignette: >
  %\VignetteIndexEntry{Integrating phylogenetic trees}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

Tree labels and data identifiers are compared exactly. A supplied tree also
defines the order of its network partition. The input trees are not rotated or
otherwise optimized, so topology and branch lengths remain under user control.

```{r}
library(ggbipartite)
```

```{r}
interaction_matrix <- matrix(
  c(
    3, 0, 1,
    1, 4, 0,
    0, 2, 5
  ),
  nrow = 3,
  byrow = TRUE,
  dimnames = list(
    c("host_a", "host_b", "host_c"),
    c("otu_a", "otu_b", "otu_c")
  )
)

row_tree <- ape::read.tree(
  text = "((host_a:1,host_b:1):1,host_c:2);"
)
column_tree <- ape::read.tree(
  text = "(otu_a:1,(otu_b:1,otu_c:1):1);"
)
```

Abundance layouts use connector panels between tree tips and boxes.

```{r}
#| fig-width: 10
#| fig-height: 5
abundance_layout <- layout_bipartite(
  interaction_matrix,
  row_tree = row_tree,
  column_tree = column_tree
)
plot_bipartite(abundance_layout)
```

Binary layouts align node centres directly with tree tips and therefore do not
need connector panels.

```{r}
#| fig-width: 9
#| fig-height: 5
binary_layout <- layout_bipartite(
  interaction_matrix,
  interaction = "binary",
  row_tree = row_tree,
  column_tree = column_tree
)
plot_bipartite(binary_layout)
```

By default, either side of an ID mismatch is an error and the message lists the
set differences. If discarding unmatched objects is scientifically justified,
opt in explicitly:

```{r}
#| eval: false
layout_bipartite(
  interaction_matrix,
  row_tree = row_tree,
  unmatched_data = "drop",
  unmatched_tree = "prune"
)
```

The composer returns a `ggbipartite_plot`, not a bare patchwork. Its named
`components` can be edited before `as_patchwork()` or printing. The network and
component data remain available through `$layout`.
